---
title: "PA1_template"
output: html_document
---

## Coursera: Reproducible Research Assignment 1

### Preparation for the Assignment

The packages used are:
```{r, echo = TRUE}
library(dplyr)
library(ggplot2)
```

Reading in the data and processing it
```{r, echo = TRUE}
activity <- read.csv("activity.csv")
str(activity)
activity$date <- as.POSIXct(activity$date, format = "%Y- %m -%d")
activity$steps <- as.double(activity$steps)
head(activity)
```

We transformed the steps to double due to replacing the missing values by their 
mean later on.

### Task1: What is mean total number of steps taken per day?

Preparing the data and use `na.rm = TRUE` to ignore missing values.
```{r, echo = TRUE}
task1 <- activity %>%
    group_by(date) %>%
    summarise(steps = sum(steps, na.rm = TRUE))
head(task1)
```

Make the plot with base R:
```{r, echo = TRUE}
hist(
    task1$steps,
    xlab = "Number of steps taken per day",
    xlim = c(0, max(task1$steps)),
    breaks = nrow(task1),
    main = "Distribution of number of steps taken per day",
    col = "turquoise3"
)
box()
```

Calculate and save the mean and the median:
```{r, echo =TRUE}
mean_task1 <- mean(task1$steps)
mean_task1

median_task1 <- median(task1$steps)
median_task1
```

So the mean is `r round(mean_task1, digit =2)` and the median is `r round(median_task1, digit = 2)`.

### Task2: What is the average daily activity pattern?

Preparing the data for the time series plot:
```{r, echo = TRUE}
task2 <- activity %>%
    group_by(interval) %>%
    summarise(steps = mean(steps, na.rm = TRUE))
head(task2)

```

Make the plot with ggplot2:
```{r, echo = TRUE}
ggplot(data = task2, aes(x = interval, y = steps)) +
    geom_line(color = 6, size = 0.5) +
    xlab("5-minute interval") +
    ylab("average number of steps taken")+
    labs(title = "time series plot of averaged steps per 5-min interval",
         color = "types")
```

Calculating the 5-minute interval that, on average across all the days in the dataset, contains the maximum number of steps

```{r, echo = TRUE}
interval_max_steps <- which.max(task2$interval)
interval_max_steps
```

So the interval we searched for is interval `r interval_max_steps`.


### Task3: Imputing missing values

The number of rows with missing values can be calculated by
```{r, echo = TRUE}
na_rows <- nrow(activity %>% filter(is.na(steps)))
na_rows
```

The method we are going to use to impute the missing values is by replacing them by the mean of the corresponding 5-minute interval.
```{r, echo = TRUE}
activity_noan <-
    left_join(activity,
              task2 %>% rename(steps_mean = steps),
              by = c("interval" = "interval")) %>%
    mutate(steps = case_when(is.na(steps) ~ steps_mean, TRUE ~ steps)) %>% 
    select(-steps_mean)
head(activity_noan)
```

For the new histogram the data is processed
```{r, echo = TRUE}
task1_noan <- activity_noan %>%
    group_by(date) %>%
    summarise(steps = sum(steps, na.rm = TRUE))
head(task1_noan)
```

And then the plot is made by the code
```{r, echo = TRUE}
hist(
    task1_noan$steps,
    xlab = "Number of steps taken per day/no NAs",
    xlim = c(0, max(task1_noan$steps)),
    breaks = nrow(task1_noan),
    main = "Distribution of number of steps taken per day",
    col = "turquoise3"
)
```

The new mean and median are
```{r, echo = TRUE}
mean_task1_noan <- mean(task1_noan$steps)
mean_task1_noan

median_task1_noan <- median(task1_noan$steps)
median_task1_noan
```

The mean with missing values which we computed in Task 1 was `r round(mean_task1, digit =2)` and the median was `r round(median_task1, digit = 2)`.
We now have `r round(mean_task1_noan, digit =2)` as the mean and `r round(median_task1_noan, digit = 2)` as the median when the missing values are replaced by their corresponding mean. Before the `NA`s were replaced they were counted as being 0 (by `na.rm = TRUE`). You can see in the histogram that the frequency of no-step-days is much lower than before imputing missing values. Mean and median are higher now.


### Task 4: Are there differences in activity patterns between weekdays and weekends?

We again prepare the data for the next analysis by adding a factor variable which tells us if a day was a weekday or not
```{r, echo = TRUE}
task4 <- activity_noan %>%
    mutate(dtype = as.factor(case_when(
        weekdays(date) %in% c("Samstag", "Sonntag") ~ "weekend",
        TRUE ~ "weekday"
    )))  
head(task4)
```

After adding the new variable we calculate the mean for each 5-minute interval and each daytype

```{r, echo = TRUE}
task4_plot <- task4 %>%
    group_by(dtype, interval) %>%
    summarise_at(vars(steps), funs(mean(.,na.rm =TRUE)))
head(task4_plot)
```

The result is
```{r, echo = TRUE}
ggplot(data = task4_plot, aes(x = interval, y = steps, color = dtype)) +
    geom_line(size = 0.5) +
    facet_wrap(~dtype ,ncol = 1, nrow = 2)+
    xlab("5-minute interval") +
    ylab("average number of steps taken")+
    labs(title = "time series plots of averaged steps per 5-min interval",
         color = "types")
```


####Save as html




